\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage[numbers]{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{hyperref}
\newcommand{\email}[1]{\href{mailto:#1}{{\UrlFont #1}}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
% \usepackage{verbatim}
% \makeatletter
% \def\verbatim@font{\sffamily\upshape}
% \makeatother

\bibliographystyle{plainnat}
\usepackage{macros}
\newcommand{\ignore}[1]{}
\newcommand{\todo}[1]{{\color{red}\textbf{\{#1\}}}}

\begin{document}

\title{The Code That Never Ran\\Modeling Attacks on Speculative Evaluation
  \thanks{Identify applicable funding agency here. If none, delete this.}
}

\author{
\IEEEauthorblockN{Craig Disselkoen}
\IEEEauthorblockA{\textit{University of California San Diego}\\
  \textit{Mozilla Research Internship}\\
  \email{cdisselk@cs.ucsd.edu}}
\and
\IEEEauthorblockN{Radha Jagadeesan}
\IEEEauthorblockA{\textit{DePaul University}\\
  \email{rjagadeesan@cs.depaul.edu}}
\and
\IEEEauthorblockN{Alan Jeffrey}
\IEEEauthorblockA{\textit{Mozilla Research}\\
  \email{ajeffrey@mozilla.com}}
\and
\IEEEauthorblockN{James Riely}
\IEEEauthorblockA{\textit{DePaul University}\\
  \email{jriely@cs.depaul.edu}}
}


\maketitle

\begin{abstract}
  This paper studies information flow caused by speculation mechanisms
  in hardware and software.  The Spectre attack shows that there are
  practical information flow attacks which use an interaction of
  dynamic security checks, speculative evaluation and cache timing.
  Previous formal models of program execution have not been designed
  to model speculative evaluation, and so do not capture attacks such
  as Spectre. In this paper, we propose a model based on pomsets which
  is designed to model speculative evaluation. The model provides a
  compositional semantics for a simple shared-memory concurrent
  language, which captures features such as data and control
  dependencies, relaxed memory and transactions. We provide models for
  existing information flow attacks based on speculative evaluation
  and transactions.  We also model new information flow attacks based on compiler
  optimizations. The new attacks are experimentally validated against
  gcc and clang.  We develop a simple temporal logic that supports 
  invariant reasoning.
\end{abstract}

\input{intro}
\input{infoflow}
\input{model}
\input{examples}
\input{tx}
\input{experiments}
\input{logic}
\input{outro}
\bibliography{bib}

\clearpage
\appendix
\input{appendix}

\end{document}
